### 2015-04-15 Dissertation/chapter-comparison_algorithms.tex
Major update from journal article polishing

\documentclass[twoside]{article}
\usepackage{ecj,palatino,epsfig,latexsym,natbib}
\usepackage{url}
\usepackage{verbatim}
\usepackage{subcaption}
\usepackage[noend]{algpseudocode}

\newcommand{\includegraphicswide}[1]
{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}

\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}

\usepackage{array}
\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}

\renewcommand{\baselinestretch}{1}

%% do not add any other page- or text-size instruction here

\parskip=0.00in

\begin{document}

% 45-character limit on header
\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}

\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
        \addr{Department of Computer Science and Engineering, Michigan State University, 
        East Lansing, 48823, United States}
\AND
       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
        \addr{Department of Computer Science and Engineering, Michigan State University, 
        East Lansing, 48823, United States}
}

\maketitle

\begin{abstract}
%Should be about 200 words
The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring any user specified parameters.
P3's primary innovation is to replace the generational model with a pyramid of
multiple populations that are iteratively created and expanded. In combination
with local search and advanced crossover P3 scales to problem difficulty, exploiting
previously learned information before adding more diversity.

Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
all five advanced comparison algorithms. This improvement includes requiring less evaluations
to find the global optimum and better fitness when using
the same number of evaluations. Using both algorithm analysis and comparison we find P3's
effectiveness is due to its ability to properly maintain, add, and exploit diversity.

Unlike the best comparison algorithms, P3 was able to achieve this quality without any
problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
sacrifice quality for applicability. Therefore we conclude that
P3 is an efficient, general, parameter-less approach to black box
optimization which is more effective than existing state-of-the-art techniques.

\end{abstract}

\begin{keywords}

Genetic algorithms, 
linkage learning,
local search,
parameter-less.

\end{keywords}

\section{Introduction}
A primary purpose of evolutionary optimization is to efficiently find good solutions
to challenging real world problems with minimal prior knowledge about the problem itself.
This driving goal has created search algorithms which can escape user bias to create
truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
While it is not possible for any algorithm to do better than random search across all possible
problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
landscape has structure and then biasing the algorithm toward exploiting that structure.

In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
biased through parameters. This can be beneficial as it allows practitioners to inject their
knowledge about the shape of the search landscape into the algorithm.
However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
domain experts may not understand how the parameters will interact with the problem or with each other.
To further complicate matters, there is mounting evidence that parameter values should change
during search~\citep{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.

There have been periodic efforts to reduce or remove the need for parameter tuning.
\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
were included in each solution's genome and themselves underwent evolution. This allowed search
to optimize some of its own parameters, resulting in a reduced need for expert tuning.
\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.

One area that has been very effective at reducing the number of algorithm parameters is
model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
both require only a single parameter: population size. \cite{posik:2011:parameterless}
leveraged model building to create a fully parameter-less algorithm, but it is restricted to
only order-k, fully decomposable, noiseless problems.

Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
This method uses a pyramid structure of populations to combine model based search with local search
to achieve parameter-less optimization. Initial results suggest that unlike
previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
parameterized search algorithms. In this work we shall: extend these results to cover more
comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.

\section{Comparison Optimizers}
\label{sec-optimizers}

@@

\subsection{$1+(\lambda, \lambda)$}
\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
the advantages of performing crossover on One Max. This comparatively simple
algorithm maintains only a single individual and a self-controlled parameter $\lambda$.

@@

\subsection{Linkage Tree Genetic Algorithm}
\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
GAs across many benchmarks. Since its introduction, many variants of LTGA have been
proposed~\citep{thierens:2011:gomea,goldman:2012:ltga} so for clarity we have chosen the version
presented by \cite{thierens:2013:ltgahiff} as our model.

@@

\section{Parameter-less Population Pyramid}
\label{sec-p3}
\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
performing optimization which does not require the user to provide any parameters. This is
achieved by combining efficient local search with the model building methods of LTGA using
an iteratively constructed hierarchy of populations.

The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
local optimum has not yet been added to any level of the pyramid, the solution is added to the lowest
population $P_0$.

Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
and the number of populations stored increases over time. Initially the pyramid contains no solutions
or populations, meaning the user does not need to specify a population size.

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$ and rebuild $P_0$'s model
      \State Add solution to $hashset$
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$
      \If{solution's fitness has improved}
        \If{solution $\notin hashset$}
          \State Add solution to $P_{i+1}$ and rebuild $P_{i+1}$'s model
          \State Add solution to $hashset$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations and $hashset$ is a set
           of all solutions in $pyramid$.}
  \label{fig-p3}
\end{figure}

To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
clusters are identified at the start of each generation and are used
to create all offspring in that generation. As P3 does not perform serial
generations, P3 instead rebuilds the model each
time a solution is added to a population. Furthermore, unlike
our chosen variant of LTGA, all solutions in the population are used to generate the model, not just the
winners of a binary tournament. We can do this because even the
worst solutions in the pyramid are already high quality due to local search.
Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
to provide no significant improvement. A likely cause was that that study applied
local search to every solution, not just the initial population, resulting in significant overhead.

Beyond the changes in population structuring, P3 modifies
LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
from \emph{last merged first} ordering to \emph{smallest first} ordering.
This method applies gene clusters during crossover based on how many genes
are in each cluster,\footnote{Ties are broken randomly.} and not on how tightly linked those genes are.
\cite{goldman:2012:ltga} found that this alternative was better at preserving
diversity, and therefore required smaller populations.

P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
Instead of choosing a single genetic donor for each cluster, P3 iterates over the
population in a random order until a solution in the population is found which
has a least one gene different for that cluster of genes from the improving solution.
This process increases the likelihood of an evaluation being performed for every cluster,
and helps test rare gene patterns in the population.

In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
gene frequency information for all $\mu$ solutions in the population. P3 does not store
a single population, and it does not have a fixed $\mu$ size for any population. However,
each time a solution is added to the population, it requires $O(N^2)$ time to update
the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
The model is then used immediately to perform up to one evaluation for each of the up
to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
rebuild the model more frequently per solution in the population, it also performs
a number of local search evaluations which are quite efficient, meaning theoretical
comparisons of their speed are difficult to perform. As a final note, P3's repeated
attempts to find a useful donation make it less likely than LTGA to skip evaluations,
but has an added cost to find these donations. Repeated donations could require as much
as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
operation actually saves more overhead than it costs by increasing the number of evaluations
per model rebuild.

Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
but for the ways in which they interact. By using the hill climber to optimize randomly
generated solutions, the underlying pairwise relationships in the problem are exposed. As
a result, detecting clusters for use by crossover is much more effective. The crossover operator
is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
strictly improve to be added to the next level of the pyramid. This is balanced by continual
integration of new random solutions. Furthermore, each random restart decreases the probability
of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
gene clusters in smallest first order during crossover as this reduces the probability of genetic
hitchhikers.

Other algorithms have proposed using multiple concurrent populations.
\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
upward. This allows for continuous integration of diversity as the lowest population is reseeded
with random solutions. However, this method resulted in increased parameterization as not only was
a population size required, but also new parameters for how frequently generations advanced between levels and
how many total levels to have. \cite{harik:1999:parameterlessga} used multiple independent
populations of different sizes as a method for removing the population size parameter, but
doing was provably less efficient than using an optimal population size as no information is shared
between the populations.

\section{Problem Descriptions}
\label{sec-problems}

\subsection{Single Instance Problems}
Understanding how a stochastic search algorithm will behave on arbitrary and complex
search landscapes can be exceedingly difficult. Therefore a common practice for
algorithm understanding is to perform search on well defined, well understood
landscapes. To be of interest these landscapes need to represent interesting
and important aspects of real world problems.

Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
to as traps.
Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
other solutions lead to a local optima of all 0s. This problem tests an algorithm's
ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
is at preserving building blocks. Any crossover event that mixes bits from different parents in
the same trap will likely result in that trap being optimized to the local optima. For our experiments
we chose $k=7$ to ensure highly deceptive traps.

\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t, &  t<k\\
       k,   &  t = k
     \end{array}
   \right.
  \label{eq-deceptive-trap}
\end{equation}

\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
Trap problem trivial. This is because local search is able to optimize each trap to one of the two
optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
In order to make the problem more difficult, the authors proposed the Deceptive
Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
of local optima in each trap.
With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
This means that half of all ways to set the trap are 1 bit local optima.
More generally, the number of local optima grows at $\Theta(2^{k-1})$.
As a result, the Deceptive Step Trap is much more challenging for linkage
learning techniques, while still being highly deceptive.
\begin{equation}
   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
  \label{eq-deceptive-step-trap}
\end{equation}

Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
and only If (HIFF) problem~\citep{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
leaf and each internal node is the subset of genes contained in its children. If all genes represented
in a node of the tree are set to the same value, they score equal to the size of the set. In this way
small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.

As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
\begin{equation}
  10n + \sum_{i=1}^{n}\left [ x_i^2-10\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
  \label{eq-rast}
\end{equation}

\subsection{Randomly Generated Problem Classes}
While well defined landscapes can provide specific insights into how an algorithm works,
their static nature can be misleading. Specifically, algorithm quality might
be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
test of an algorithm's black box effectiveness is to work with randomly generated instances
drawn from a problem class. When tested over a sufficiently large sample it is then possible
to draw more general conclusions about the algorithms effectiveness. The challenge with these
landscapes is determining the global optimum to gauge if an algorithm was successful.

Perhaps the most common model for generating random rugged landscapes is the NK model. An
NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
other genes in the genome. This fitness is specified using a randomly generated table of
fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
if epistasis is set such that each gene depends on the $K$ directly following it
in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.

\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
to either -1 or 1, and vertex values must be -1 or 1.

As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}
While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each problem size.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

@@

\section{Finding the Global Optimum}
\label{sec-optimum}

\begin{figure}
  \begin{center}
  \includegraphicsfit{evals-to-success}
  \end{center}
  \caption{Comparison of the median number of evaluations to reach the global optimum for
           the six different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no data element
           is shown.  Results given on a log-log scale.}
  \label{fig-evals-to-success}
\end{figure}

Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
algorithms to find the global optimum for multiple sizes of each problem. Each data point in
Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
are treated as requiring more evaluations than any successful run. If the median run was not successful
no point is shown.
Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
between techniques with different success rates.
The maximum problem size used for each problem was set to be the largest size we could
feasibly determine LTGA's optimal population size. For many larger problems results are not shown
for hBOA due to the extreme computational cost required to optimally set the population size.

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
according to the binomial test.
Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
$p < 10^{-5}$ for all problems.

\subsection{Local Search}
The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes. This is especially true for the three randomly generated problem classes. These problems
may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
better than any other algorithm. We believe this is
because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps.
This ability also leads $1+(\lambda, \lambda)$
to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.

On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
Only on MAX-SAT are these optimizers competitive
at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
was an order of magnitude smaller than the largest size tested for most other problems. As the problem
size increases the number of local optima increases exponentially, which explains why Random Restart Hill
Climbing was unable to scale.
For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
of changes required to reach the global optimum. This
behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.

\subsection{Model Building}
Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
unlike the single instance problems, gene epistasis in the randomly generated
problem classes cannot be be perfectly represented with a linkage tree.

Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
their results are on HIFF.
However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
lower order subproblems. Therefore both methods require one generation per subproblem order.

\subsection{P3}
Unlike the other model based methods, P3 generally outperforms
both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
problems and model building to solve harder ones.

Furthermore, the interaction between these two optimization tools explains some of the reason P3
outperforms each method alone.
On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
linkage detection and high quality donation.
On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.

\section{Fitness Over Time}
\label{sec-overtime}
\begin{figure}
  \begin{center}
  \includegraphicsfit{fitness-over-time}
  \end{center}
  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
  \label{fig-fitness-over-time}
\end{figure}

For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
is set to include the slowest P3 run to reach the global optimum.

% 181 total
% 121 has P3 best
%   9 hc
%  50 lambdalambda
%  20 hboa
%  18 phboa
%  27 LTGA

\subsection{Quantitative Comparison}
Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
never outperformed P3, doing so only 9 times.
The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
according to the binomial test.

\subsection{Local Search}
Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
this method is slow in reaching the global optima in many of these problems which causes it to eventually
be overtaken by the model building techniques.

\subsection{Model Building}
Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
In hBOA this is taken to the extreme, with all fitness
improvement made at the very end of search. In both cases this is caused by model building. Before the model
is accurate little improvement is made. Once it is accurate, fitness improves dramatically.

At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
to improve.
Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.

On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
and improvement to global optimum. The early period corresponds with
initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
Subsequent generations then make only minor fitness improvements. Once the model
becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
enters a second period of rapid improvement.

\begin{comment}
This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
\end{comment}

\subsection{P3}
The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
least as high as Hill Climbing. As such P3 is better than a simple
hill climber regardless of how long each technique is run and irrespective of how high quality the solution
found has to be.

Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
P3 is able to immediately integrate optimal versions of each
trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.

\section{Computational Expenses}
While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
complexity, in some domains this will not be true. Model based methods are especially likely to violate
this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
we provide data on both its algorithmic complexity and wall clock time.

\subsection{Operation Counting}
\begin{figure}[t]
  \begin{centering}
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{rebuilds}
      \end{centering}
      \caption{Rebuilds}
      \label{fig-rebuilds}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{donations}
      \end{centering}
      \caption{Donations}
      \label{fig-donations}
    \end{subfigure}
  \end{centering}
  \caption{Estimated computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
  \label{fig-costs}
\end{figure}

When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
provides some insight into how often these two aspects of the algorithm are utilized.

Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
In order to calculate this value we recorded how many times search rebuilt the model during each run.
Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
If the cost of model building
scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
no more than twice the algorithmic cost of an evaluation even using 2048 bits.

When applying a crossover subset, P3 tries random donors from the population until one is found with at least
one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
important note is that each donation may range in size from a single bit up to $N-1$. However,
repeated donation attempts are far more likely to happen with smaller clusters.
As such this may cause some super-linear growth
in P3, but its unlikely to be very high.

\subsection{Wall Clock Performance}
\begin{figure}
  \begin{center}
  \includegraphicsfit{seconds-to-success}
  \end{center}
  \caption{Comparison of the median number of seconds to reach the global optimum for
           the six different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no data element
           is shown.  Results given on a log-log scale.}
  \label{fig-seconds-to-success}
\end{figure}

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
of 100 runs, with unsuccessful runs treated as slower than successful runs.
\begin{comment}
In total these runs
took over 6 computing years to complete. We were therefore forced to use
diverse hardware potentially running multiple processes in parallel.
As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. However, due to
the size of the dataset and the general regularity in the results some general conclusions
can still be made.
\end{comment}
These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.

>>>>

\chapter{Comparison Optimizers}
\label{chap-optimizers}

@@

%%%%%%%%%%%% Lambda Lambda %%%%%%%%%%%%%%%%
\section{$1+(\lambda, \lambda)$}
\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
the advantages of performing crossover on One Max. This comparatively simple
algorithm maintains only a single individual and a self-controlled parameter $\lambda$.

@@

\section{Linkage Tree Genetic Algorithm}
\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
GAs across many benchmarks. Since its introduction, many variants of LTGA have been
proposed~\citep{thierens:2011:gomea,goldman:2012:ltga} so for clarity we have chosen the version
presented by \cite{thierens:2013:ltgahiff} as our model.

@@

%%%%%%%%%%%%%%%% Tuning %%%%%%%%%%%%%%%%%%%%%
\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}

@@

While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each problem size.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@
